Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
Appendix wip

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\Char}{\textbf{Char}}
\newcommand{\stringg}{\texttt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
% %
%     \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
% %
%     \and
% %
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
    \and
    %
    \inferrule
    {~}
    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
    %
    \and
    %
    \inferrule
    {~}
    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e : A}
\end{mathpar}
  \caption{Judgmental equality for linear terms}
  \label{fig:jdgeq}
\end{figure}
\steven{Do we need inverse for $\ltonl A$ in judgmental equality figure?}
\steven{Do we need to say anything about the equalizer proofs being unique in
  judgmental equality figure?}

@@

\section{Denotational Semantics}

\begin{definition}
  We define the unit grammar
  \[ I\,w = \{ ()\pipe w = \varepsilon \}\]
  and given grammars $A$ and $B$ we define their tensor product
  \[ (A \otimes B)\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A\,w_1 \wedge b \in B\,w_2 \} \]
  This extends to a monoidal structure on the category of grammars. Note that we defined in the semantics of linear types $\sem{I}\gamma = I$ and $\sem{A \otimes B}\gamma = \sem{A}\gamma \otimes \sem{B} \gamma$
\end{definition}

@@

We define denotations of linear terms in
\Cref{fig:linear-term-semantics}. Note that the denotations interpret
typing derivations, not raw terms, as the data of how contexts are
split is needed in order to construct the correct associator
functions.
\max{todo: describe associators, define denotations, prove axioms}
\begin{figure}
  \begin{mathpar}
  \inferrule
  {\Delta_1 \vdash e_1 : A_1 \and \Delta_2 \vdash e_2 : A_2}
  {\sem{(e_1,e_2)}\gamma\,w\,\delta
    = \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}
  }

  \inferrule
  {\Delta_2 \vdash e : A \otimes B \and \Delta_1,a:A,b:B,\Delta_3 \vdash e' : C}
  {\sem{\letin{(a,b)}{e}{e'}}\gamma\,w\,\delta =
    \letin{(w_1,w_2w_3,\delta_1,(w_2,w_3,\delta_2,\delta_3))}{m_{\Delta_1,\Delta_2,\Delta_3}(\delta)}
    \letin{(w_{2,1},w_{2,2},a,b)}{\sem{e}\gamma\,w_2\,\delta_2}
    {\sem{e'}\gamma\,w\,(w_1,w_{2,1}w_{2,2}w_3, \delta_1, (w_{2,1},w_{2,2}w_3, a, (w_{2,2},w_3, b, \delta_3)))}
    }
\end{mathpar}
  \caption{Denotations of linear terms}
\end{figure}

\begin{definition}
  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
  \sem{\Delta_2}\gamma$.

  This can be extended to a sequence of contexts of any length.
\end{definition}

\begin{lemma}[Equations for $\otimes$]

\end{lemma}

\begin{lemma}[Equations for $I$]

\end{lemma}

\begin{lemma}[Equations for $\lto$]

\end{lemma}

\begin{lemma}[Equations for $\tol$]

\end{lemma}

\begin{align*}
  \sem{(e_1,e_2)}\gamma\,w\,\delta &=
  \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}\\
\end{align*}

>>>>

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\semg}[1]{\sem{#1}\gamma}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\Char}{\textbf{Char}}
\newcommand{\stringg}{\texttt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\semcat}{\mathbf{C}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

    % \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e {e'} a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : B \tol A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {M} \equiv \subst {e} M x : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
% %
%     \and
%
%     \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
% %
%     \and
% %
%     \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} e a : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} {e \otimes e'} e'' \equiv e'' \{ e/a , e'/b \} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} e c : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(M, e)} {e'} \equiv e' \{ M/x , e/a \} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
    \and
    %
    \inferrule
    {~}
    {\Gamma ; \Delta \vdash \equalizerpi {\equalizerin {e}} \equiv e : A}
    %
    \and
    %
    \inferrule
    {~}
    {\Gamma ; \Delta \vdash \equalizerin {\equalizerpi {e}} \equiv e :
      \equalizer e f g}
\end{mathpar}
  \caption{Judgmental equality for linear terms}
  \label{fig:jdgeq}
\end{figure}
\steven{Do we need inverse for $\ltonl A$ in judgmental equality figure?}
\steven{Do we need to say anything about the equalizer proofs being unique in
  judgmental equality figure?}

@@

\section{Denotational Semantics}
\steven{Define a Kleene, Lambek, and Chomsky category}

Let $\semcat$ be a Chomsky category.

\steven{Finish the denotation of lin types figure for spf, el , map , mu}
\begin{figure}
  \begin{align*}
    \semg{I} &= I \\
    \semg{A \otimes B} &= \semg{A} \otimes \semg{B} \\
    \semg{A \lto B}  &= \semg{A} \lto \semg{B} \\
    \semg{B \tol A} &= \semg{B} \tol \semg{A} \\
    \semg{\bigoplus_{x:X} A} &= \coprod_{x:X} \semg{A} \\
    \semg{\bigwith_{x:X} A} &= \prod_{x : X} \semg{A} \\
    \semg{\uparrow A} &= Hom(I,A) \\
    \semg{\equalizer{a:A}{f}{g}} &= Eq\left( \semg{f} , \semg{g} \right) \\
    \sem{L}\gamma &= \semcat_0\\
    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
  \end{align*}
  \caption{Denotation of Linear Types}
\end{figure}

@@

\begin{definition}[Denotation of Linear Substitutions]
  The semantics of a linear substitution
  $\delta : \textrm{Subst}(\Delta' , \Delta)$ are given as maps
  $\semg{\delta} : \semg{\Delta'} \to {\Delta}$. Define $\semg{\delta}$ by
  recursion on $\delta$:
  \begin{align*}
    \semg{\cdot} &= id_{I} \\
    \semg{\delta,e/a} &= \semg{\delta} \otimes \semg{e} \circ m_{\Delta_{1},\Delta_{2}}
  \end{align*}
where $\Delta_{2} \vdash e : A$ and $\Delta' = \Delta_{1},\Delta_{2}$.
\end{definition}

\begin{theorem}
  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
  \sem{\Gamma}$ there is a natural isomorphism $m_{\Delta_1,\Delta_2} :
  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
  \sem{\Delta_2}\gamma$.

  This can be extended to a sequence of contexts of any length.
\end{theorem}
\begin{proof}
Construct $m_{\Delta_{1},\Delta_{2}}$ by recursion on $\Delta_{2}$.
\begin{align*}
  m_{\Delta_{1},\cdot} & = \rho^{-1} \\
  m_{\Delta_{1},(\Delta_{2}, a : A)} & = \alpha \circ m_{\Delta_{1},\Delta_{2}} \otimes id
\end{align*}
\end{proof}

\begin{lemma}
  \label{lem:subst}
  For each term $\Delta \vdash e : A$ and substitution
  $\delta : \textrm{Subst}(\Delta', \Delta)$, the semantics of $\delta$ acting
  on $e$ splits into the composition $\semg{e[\delta]} = \semg{e} \circ \semg{\delta}$.
\end{lemma}

\subsection{Grammar Semantics for Linear Terms}
Here we define denotations of linear terms. Note that the denotations interpret
typing derivations, not raw terms, as the data of how contexts are
split is needed in order to construct the correct associator
functions. Further, we demonstrate that the denotational semantics resprects the
equational theory of \theoryabbv.

\subsubsection{Variables}
Note that the denotation of a singleton context $a : A$ is given as
\[
  \semg{a : A} = \semg{\cdot , a : A} = I \otimes \semg{A}
\]

So for $a: A \vdash a : A$, the denotation of a single variable term
$\semg{a} : \semg{a : A} \to \semg{A}$ is given by the left unitor
\[
  \semg{a} = \lambda
\]

\subsubsection{Linear Unit}
% I
\paragraph{$I$-Introduction}
$\semg{()} : \semg{\cdot} \to \semg{I}$.
\[
\semg{()} = id_{I}
\]

\paragraph{$I$-Elimination}
$\semg{\letin {()} e e'} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$
defined in the following diagram,

\begin{center}
\begin{tikzcd}
  \semg{\Delta'_1,\Delta,\Delta'_2} \arrow[r , "m_{(\Delta'_1,\Delta),\Delta'_2}"] &
  \semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}
    \arrow[r , "m_{\Delta'_1,\Delta} \otimes id"] &
  \left(\semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2}
    \arrow[d , "(id \otimes \semg{e}) \otimes id"] \\
    \semg{\Delta'_1,\Delta'_2} \arrow[d , "\semg{e'}"]
  & \arrow[l , "m^{-1}_{\Delta'_1,\Delta'_2}"]
    \semg{\Delta'_1} \otimes \semg{\Delta'_2}
  & \left(\semg{\Delta'_1} \otimes I \right) \otimes \semg{\Delta'_2}
  \arrow[l , "\rho \otimes id"] \\
  \semg{C}
\end{tikzcd}
\end{center}

We demonstrate that the denotations of the introduction and elimination forms
for $I$ obey the $\beta$ and $\eta$ equalities for $I$.

\paragraph{$I\beta$}
Given
$\Delta'_{1},\cdot,\Delta'_{2} \vdash e' : C$, the desired
$\beta$ law is
\[
  \semg{\letin {()} {()} {e'}} = \semg{e}
\]
\begin{proof}
\begin{align*}
  \semg{\letin {()} {()} {e'}}
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{()}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{(\Delta'_{1},\cdot),\Delta'_{2}} \\
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes id) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ \rho^{-1} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'}
\end{align*}
\end{proof}

\paragraph{$I\eta$}
Similarly, for $\Delta_{1}, a : A, \Delta_{3} \vdash e' : C$ and
$\Delta_{2} \vdash e : I$ the desired $\eta$ law is
\[
\semg{\letin {()} {e} {e'[()/a]}} = \semg{e'[e/a]}
\]
However, through application of \cref{lem:subst} it suffices to handle the case
where $e$ is a variable $a'$. That is,
\begin{align*}
\letin {()} {e} {e'[()/a]} &= (\letin {()} {a'} {e'[()/a]})[e/a'] \\
e'[e/a] &= e'[a'/a][e/a]
\end{align*}
so without loss of generality we may take may take $e$ to be variable $a'$.

\begin{proof}
\begin{align*}
  \semg{\letin {()} {a'} {e'[()/a]}}
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \semg{a'}) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'} \circ m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \lambda) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}} \\
  & = \semg{e'}
\end{align*}

\steven{Define generalized associators}

Because
$m^{-1}_{\Delta'_{1},\Delta'_{2}} \circ \rho \otimes id \circ (id \otimes \lambda) \otimes id \circ m_{\Delta_{1},\cdot} \otimes id \circ m_{\Delta'_{1},\Delta'_{2}}$
is a composition of generalized associators from
$\semg{\Delta'_{1},\Delta'_{2}}$ to itself, it is equal to the identity by the coherence theorem for monoidal categories.

Further by \cref{lem:subst},
\begin{align*}
  \semg{e'[a'/a]}
  &= \semg{e'} \circ \semg{a'/a} \\
  &= \semg{e'}
\end{align*}
Again by the coherence theorem, $\semg{a'a} = id$. Thus, $\eta$ law for $I$
holds in the denotational semantics.
\end{proof}

%tensor
\subsubsection{Tensor}
\paragraph{$\otimes$-Introduction}
$\semg{(e_{1}, e_{2})} : \semg{\Delta,\Delta'} \to \semg{A \otimes B}$ is given
by the diagram

\begin{tikzcd}
  \semg{\Delta,\Delta'} \arrow[r, "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r, "\semg{e_1} \otimes \semg{e_2}"] &
  \semg{A} \otimes \semg{G}
\end{tikzcd}

\paragraph{$\otimes$-Elimination}
$\semg{\letin {(a,b)} e f} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$
defined via the diagram,

\begin{tikzcd}
  \semg{\Delta'_1,\Delta,\Delta'_2} \arrow[r , "m_{(\Delta'_1,\Delta),\Delta'_2}"] &
  \semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}
    \arrow[r , "m_{\Delta'_1,\Delta} \otimes id"] &
  \left(\semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2}
    \arrow[d , "(id \otimes \semg{e}) \otimes id"] \\
  \semg{c} & \arrow[l , "\semg{f}"]
  \left( \left( \semg{\Delta'_1} \otimes \semg{A} \right) \otimes \semg{B}\right) \otimes \semg{\Delta'_2}
  & \left(\semg{\Delta'_1} \otimes \left( \semg{A} \otimes \semg{B}\right) \right) \otimes \semg{\Delta'_2}
    \arrow[l , "\alpha \otimes id"]
\end{tikzcd}

\paragraph{$\otimes\beta$}

\paragraph{$\otimes\eta$}

% linear function
In the semantics $\semg{- \otimes A}$ and $\semg{A \lto -}$ are left adjoint, so
for all $A$ and $B$ we have the natural isomorphism
$\phi : Hom(\semg{\Delta} \otimes \semg{A}, \semg{B}) \to Hom(\semg{\Delta}, \semg{A \lto B})$.
We use this adjunction to provide the denotation $\semg{\lamblto a e} : \semg{\Delta} \to \semg{A \lto B}$.

\[
\semg{\lamblto a e} = \phi\left( \semg{e} \right)
\]

The denotation for application of a $\lto$-function,
$\semg{\applto {e'} {e}} : \semg{\Delta,\Delta'} \to \semg{B}$ defined by the diagram

\begin{tikzcd}
  \semg{\Delta,\Delta'} \arrow[r , "m_{\Delta,\Delta'}"] &
  \semg{\Delta} \otimes \semg{\Delta'} \arrow[r , "id \otimes \semg{e}"] &
  \semg{\Delta} \otimes \semg{A} \arrow[r , "\phi^{-1}\left( \semg{e'} \right)"]
  &
  \semg{B}
\end{tikzcd}

$\semg{\lambtol a e} : \semg{\Delta} \to \semg{B \tol A}$ defined using the
natural isomorphism
$\psi : Hom(\semg{A} \otimes \semg{\Delta}, \semg{B}) \to Hom(\semg{\Delta}, \semg{B \tol A})$
acting on the following diagram

\begin{tikzcd}
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\lambda^{-1} \otimes id"] &
  \semg{a : A} \otimes \semg{\Delta} \arrow[r , "m^{-1}_{a : A , \Delta}"] &
  \semg{a : A , \Delta} \arrow[r , "\semg{e}"] &
  \semg{B}
\end{tikzcd}

The application of a $\tol$-function,
$\semg{\apptol {e} {e'}} : \semg{\Delta',\Delta} \to \semg{B}$ defined by the diagram

\begin{tikzcd}
  \semg{\Delta',\Delta} \arrow[r , "m_{\Delta',\Delta}"] &
  \semg{\Delta'} \otimes \semg{\Delta} \arrow[r , "\semg{e'} \otimes id"] &
  \semg{A} \otimes \semg{\Delta} \arrow[r , "\phi^{-1}\left( \semg{e} \right)"] &
  \semg{B}
\end{tikzcd}

% Products
Product introduction,
$\semg{\dlamb x e} : \semg{\Delta} \to \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}$ is
defined by the universal property of the product
\[
\semg{\dlamb x e} = \left( \sem{e}(\gamma , x) \right)_{(x : X)}
\]

Product elimination, $\semg{e.\pi~M} : \semg{\Delta} \to \sem{A}(\gamma , M)$

\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
  \sem{A}(\gamma , M)
\end{tikzcd}

% Sums
Sum introduction,
$\semg{\sigma~M~e} : \semg{\Delta} \to \LinSigTyLimit {x} {X} \sem{A}(\gamma, x)$

\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \sem{A}(\gamma, M) \arrow[r , "i_M"] &
  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}
\end{tikzcd}

Sum elimination,
$\semg{\letin {\sigma~x~a} {e} {e'}} : \semg{\Delta'_{1},\Delta,\Delta'_{2}} \to \semg{C}$

% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAB124YBbAc2CcAIjAY46AcgD6ARlLDR4+exFjJUgEwBfEFtLpMufIRRzK1es1aIOXXgIVrpcx+K0ACThDw94nu-yCKorq2rr6IBjYeAREcjJUtIwsbJwMMABmOAAU-tyBruoyHl4+fpz5DsFqJewATlh8ABY4AJT+3li+cHn2Qari0mF6BtHGcaQaiZYpNpwAMlhgAMqNACoAnos8uO7AAB4ewAAaRxW8-SHOpHTuiO4AgsoDoVrZnHx0PDy3pO77rR0I0ihhiJhIk2myWstkWK3WWy6uwOR1OezSmRyvR4lycsjeHy+P3cfwBHTKcHYYHcNNp5xxDwJ7E+31+-3anAazTa5K65QCuMGmiZLOJpMB4VGRliKDICQs0NS7HSWVy9KqL2ctU63X8cNWfE222Rhz2aKC+WAjPezKJbIBtS5LQ57B1-MqgtekpBYxlyDkAGYoVYlZaAMIiu3i71RaXgshBhUhuYBAQR3SJGBQPjwIigDJ1CA8JBkEA4CBIORJZMgHhSYA2zWyZ4hVotvFhagMOgAI1EAAVQeMbCqcN6C0XK9Ry0gNEnZrX64VroUtKU+T0sFAQF3ewOhzKQKPx4Xi4g52WK4gAKzzmHZLe83Xq4AwLQut2b7e7vsMQe+kwj0xE9J0QANpyvAA2O82GQbFXwkSNWRJdl3Aoetsn2O53GOCUf33AC2GPYEJzPAAWCCkHA6sFx4AA9YAAFpigw5dmzoRAHjbNjOyPPc-wPQDiIiUikFvS8kAomiYR3Pjf3-OMiOAkjTyQaCJMQAB2LQKC0IA
\begin{tikzcd}[column sep =large]
{\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
{\left( \semg{\Delta'_1} \otimes \LinSigTyLimit {x} {X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, ""]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
{\LinSigTyLimit {x} {X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "{m^{-1}_{(\Delta'_1,a:A),\Delta'_2}}"] & {\LinSigTyLimit {x} {X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
\semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
\end{tikzcd}

%equalizer
Equalizer introduction,
$\semg{\equalizerin{e}} : \semg{\Delta} \to \semg{\equalizer e f g}$ where
$\semg{e} : \semg{\Delta} \to \semg{A}$ and
$\semg{f} \circ \semg{e} = \semg{g} \circ \semg{e}$. By the universal property
of the equalizer the preceding equality induces a unique morphism
$\semg{\Delta} \to Eq(\semg{f} , \semg{g}) = \semg{\equalizer e f g}$. Define
$\semg{\equalizerin{e}}$ to be this map.

Equalizer elimination, $\semg{\equalizerpi e} : \semg{\Delta} \to \semg{A}$

\[
  \semg{\equalizerpi e} = \pi_{eq} \circ \semg{e}
\]
